---
title: 'lakemorpho: Calculating lake morphometry metrics in R'
author:  
- name: Jeffrey W. Hollister
  affilnum: '1'
affiliation:
- affilnum: 1
  affil: US Environmental Protection Agency, Office of Research and Development, National Health and Environmental Effects Research Laboratory, Atlantic Ecology Division, 27 Tarzwell Drive  Narragansett, RI, 02882, USA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: no
    template: components/manuscript.latex
  html_document: null
  word_document: null
fontsize: 11pt
capsize: normalsize
csl: components/plos.csl
documentclass: article
spacing: doublespacing
linenumbers: yes
bibliography: components/manuscript.bib
abstract: no
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{lakemorpho: Calculating lake morphometry metrics in R}
-->

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
library("knitr")
library("lakemorpho")
library("rgdal")
library("sp")
library("rgeos")
library("knitr")

opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

# Table Captions from @DeanK on http://stackoverflow.com/questions/15258233/using-table-caption-on-r-markdown-file-using-knitr-to-use-in-pandoc-to-convert-t
#Figure captions are handled by LaTeX

knit_hooks$set(tab.cap = function(before, options, envir) {
                  if(!before) { 
                    paste('\n\n:', options$tab.cap, sep='') 
                  }
                })
default_output_hook = knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  if (is.null(options$tab.cap) == FALSE) {
    x
  } else
    default_output_hook(x,options)
})
```

```{r analysis , include=FALSE, echo=FALSE, cache=FALSE}
#All analysis in here, that way all bits of the paper have access to the final objects
#Place tables and figures and numerical results where they need to go.
```

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->

\singlespace

\vspace{2mm}\hrule

Metrics describing the shape and size of lakes, known as lake morphometry metrics, are important for any limnological study.  In cases where a lake has long been the subject of study these data are often already collected and the data is openly available.  Many other lakes have this data collected, but access to the data is challenging as it is often stored on individual computers (or worse in filing cabinets) and is available only to the primary investigators.  The vast majority of lakes fall into a third category in which the data is not available.  This makes broad scale modelling of lake ecology a challenge as some of the key information about in-lake processes are unavailable.  While this valuable *in situ* information may be difficult to obtain, several national datasets exist that may be used to model and estimate lake morphometry.  In particular digital elevation models and hydrography have been shown to be predicitive of several lake morphometry metrics.  The R package `lakemorpho` has been developed to utilize this data and estimate the following morphometry metrics: surface area, shoreline length, shoreline development, maximum depth, mean depth, volume, maximum lake length, mean lake width, maximum lake width, and fetch. In this software note I describe the motivation behind developing `lakemorpho`, discuss the implementation in R, and describe the use of `lakemorpho` with an example of a typical use case.    

\vspace{3mm}\hrule
\doublespace

#Introduction

The study and quantificaiton of lake shape (i.e. lake morphology and morphometry) is one of the foundations of limnology and for students of limnology, some of the first lessons are centered around a typical suite of metrics and how to calculate them [e.g. @wetzel2001limnology].  It is also widely accepted that the morphometry of lakes and ponds can impact available nutrients and thus overall productivity.  For instance, the widely used Vollenweider input-output models that are used to estimate nutrient concentrations rely on hydraulic residence time and sometimes mean depth, both of which are derived from total lake volume [@vollenweider1975input; @milstead2013estimating]. Also, clear water versus turbid water states in lakes have been linked in part to lake morphometry, in particular mean depth [@genkai2005eutrophication].  In short, limnologists have long recognized the importance of lake morphology as one factor controlling a variety of ecological processes in lakes.

Traditional methods for calculating lake morphometry metrics have relied upon the use of paper bathymetry maps, planimeters, or unnecessary assumptions [@kalff2002limnology; @welch1935limnology; @wetzel2000limnological; @hollister2010volume]. In addition, detailed bathymetry is a requirement for most of these methods, but it's collection is only available for a relatively small number of lakes.  This is not a problem when the focus of a study is a single lake, a small number of lakes or on well studied lakes.  Relying on complete bathymetry becomes a limitation when attempting to conduct regional or national studies of lakes as bathymetry is at best difficult to find or does not exist for all lakes of interest.  In these cases alternative approaches for estimating lake morphometry are required.  

Recent work has demonstrated the ability to estimate many of these metrics from ubiquitous spatial data.  For instance, maximum depth and lake volume may be predicted using the lake polygon and surrounding topography [@hollister2010volume; @hollister2011predicting] provided by the National Hydrography Dataset Plus and the National Elevation Dataset, respectively [e.g. @usepa2005national;@gesch2009national]. The initial development of these tools were developed with proprietary tools thus limiting their use.  In an effort to reach a broader audience the tools were converted to R, expanded to include a more complete suite of lake morphometry metrics and compiled into an R Package.

#Implementation and Use in R

Using R as a Geographic Information System is now possible as several packages provide spatial data handling, geospatial analysis, and visualization.  It is becuase of these pacakges that `lakemorpho` was  implemented as an R package.  In particular,`lakemorpho` relies on the following packages: `maptools`, `rgdal`, `raster`, `rgeos`, `sp`, `geosphere`[@bivand2014maptools;@bivand2014rgdal;@bivand2014rgeos;@bivand2008applied;@pebesma2005sp;@hijmans2014raster;@hijmans2014geosphere].  In addition to these packages two external libraries, the Geospatial Data Abstraction Library (GDAL) and Geometry Engine, Open Source(GEOS), are needed.  Their availability to R and installation varies by operating system [@GDAL2012;@GEOS2013].  

##Using `lakemorpho`

Included in `lakemorpho` are, one function to create a `lakeMorpho` object,  eleven functions to caclulate morphometry metrics, a default plotting function, two example datasets, and an example `lakeMorpho` object.

A typical workflow for using `lakemorpho` to calculate lake metrics would include
pulling spatial data into R (e.g. as shapefiles, tiff, etc.), creating a lakemorpho object  and calculating the desired lake morphometry metrics.  The following sections provide details on the type of input data required and dicsuss the use of the functions, inlcluding examples with the provided example data.

###The `lakeMorpho` Class and `lakeSurroundTopo`  
Many of the lake morphometry metrics rely on the same information about the lake.  For instance, the functions to estimate maximum depth, mean depth, and volume rely on statistical summaries of the surrouding topography as well as the maximum in-lake distance to shoreline.  [@hollister2011predicting; @hollister2010volume]. To avoid recaluclating these values, a `lakeMorpho` class was created to store the information on surrounding topography as well as the original datasets.  This object is required input for all of the lake morphometry functions in the the `lakemorpho` package.  In addition to this, an object of class `lakeMorpho` also holds the initial datasets and, optionally, can store the spatial objects that result from some of the lake morphometry functions.  At a minimum, a `lakeMorpho` object contains.

- "lake" - A `SpatialPolygons` or `SpatialPolygonsDataFrame` object of the originally input lake data.
- "elev" - A `RasterLayer` representing the elevation in a suitably large area around the lake.
- "surround" - A `SpatialPolygons` or `SpatialPolygonsDataFrame` object representing the land area defined as the surrounding topography.
- "lakeDistance" - A `RasterLayer` object of the euclidean distance from the shoreline to center of each pixel.  Maximum value is equal to the maximum in-lake distance.
- "lakeOnEdge" - A logical value indicating if the `lakeMorpho` value "surround" is on the edge of the value "elev". 

The `lakeSurroundTopo` function is the primary mechanism for creating a `lakeMorpho` object.  There are two required inputs and one optional input for `lakeSurroundTopo`.  The first required input is a `SpatialPolygons` or `SpatialPolygonsDataFrame` of the lake [@pebesma2005sp].  Only a single lake is accepted as input, although this lake may be composed of multiple polygons (i.e. a lake with islands).  If metrics for multiple lakes are required they will need to be passed to the suite of `lakemorpho` functions separately.  The second required input is a `RasterLayer` of the elevation surrounding the lake [@hijmans2014raster].  The default raster size is taken from the resolution of the input elevation data but may be specified separately.  The third input specifies the area representing the surrounding topography.  By default this is a buffer of the lake shoreline with the buffer width equal to the maximum in-lake distance.  An optional `SpatialPolygons` object of any polygon intersecting the lake (e.g. catchements) can be used to define the surrounding topography instead of the default buffer.  An object of class `lakeMorpho` is returned from `lakeSurroundTopo`

In addition to providing accepted inputs, users should pay attention to both the extent of the input elevation dataset as well as the coordinate reference systems used.  First, the elevation data must be of a large enough extent so that the surrounding topography does not inlcude land area outside that extent (i.e would return NA values).  As noted above, the `lakeOnEdge` item indicates if the surrounding topography is on the edge of the input elevation and thus returns NA values.  Second, all of the functions of `lakemorpho` assume that projections have been handled prior to creating the `lakemorpho` class or calculating the metrics.  If the input data are not of the same projection, `lakeSurroundTopo` will return an error.  The data must be reprojected into the same coordinate reference system (CRS).  Care must be taken in choosing a CRS as area and length measurments will vary between different CRS. 

Usage of `lakeSurroundTopo` and generating a `lakeMorpho` object from the example data included with `lakemorpho` is done as follows:

```{r lakeSurroundTopo_example}
#Load data
data(lakes)

#Create lakeMorpho object, example_lakeMorpho, with required inputs
example_lakeMorpho <- lakeSurroundTopo(exampleLake, exampleElev)
```

The resulting object contains the minimum set of values that are all of the expected class.

```{r lakeSurroundTopo_exmaple_output}
lapply(example_lakeMorpho,class)
```


###Lake Morphometry Functions
Each of the remaining functions all expect a `lakeMorpho` object as input and all return a numeric value.  Some of the functions do have a side effect of adding a spatial object to the input `lakeMorpho` object.  

####lakeFetch	
Fetch is the maximum open water distance in a given direction and can be used an indicator of mixing as greater fetch implies greater potential for waves[NEED REF]. The lakeFetch() function calculates fetch along an input bearing.  The input bearing may be any value from 0 to 360 where 0 and 360 both represent north, although the fetch for opposite directions (e.g. east and west) are identical.

To calulcate the fetch of an input lake use:

```{r lakeFetchExample}
#Fetch for North
lakeFetch(example_lakeMorpho, 0)
lakeFetch(example_lakeMorpho, 360)

#Fetch for West
lakeFetch(example_lakeMorpho, 270)
```


####lakeMaxDepth	
Estimate maximum lake depth

####lakeMaxLength
Caluclate maximum lake length

####lakeMaxWidth
Function to find line representing maximum lake Width

####lakeMeanDepth	
Function to return lake Mean Depth

####lakeMeanWidth	
Function to return lake Mean Width

####lakeShorelineDevelopment	
Function to calculate shoreline development

####lakeShorelineLength	
Function to calculate shoreline length

####lakeSurfaceArea	
Return lake surface area

####lakeVolume
Caluclates Lake Volume in R

####calcLakeMetrics  
Calculate all Lake Morphometry Metrics
 

#Software Availability

The `lakemorpho` version 1.0.1 package is currently available directly from the Comprehensive R Archive Network (CRAN) and may simply be installed and loaded in R via:

```{r, eval=FALSE}
install.packages('lakemorpho')
library('lakemorpho')
```

To access the help pages (including a version of this manuscript) use.

```{r, eval=FALSE} 
help(package='lakemorpho')
```

There are tentative plans to continue developing new functions for `lakemorpho` and these new features will be available first through the development version on GitHub at [http://github.com/usepa/lakemorpho](http://github.com/USEPA/lakemorpho).  To install and load the development version requires use of the `devtools` package.  This may be done with:

```{r, eval=FALSE}
install.packages('devtools')
library('devtools')
install_github('USEPA/lakemorpho')
library(lakemorpho)
```

#References